---
title: "US Presidential 2016 - California"
output:
  html_document:
    theme: united
    toc: yes
  pdf_document:
    toc: yes
---

by E. Rabaute 



### I) Objectives

The objective of this analysis is to provide valuable information about 2016 US presidential campaign donation using data analysis methods and tools.

RStudio will be used for data wrangling, and data explonary analysis.

This project is the last task validating Udacity explonary analysis MOOC.

### II) Data analysed

#### a) Format and volume
The data used for this analysis is a csv file downloaded from http://classic.fec.gov/  related from campaign Finance in Calfornia state.

Data size: ~250 MB
Type: CSV file (comma separation)

Rows:  1 304 346

In order to accelerate computation for exploratory analysis, I'd sample all the data around 10%. Then, once all code and plots were generated, I computed all data.


```{r echo=FALSE, message=FALSE, warning=FALSE}
library(ggmap)
library(ggplot2)
library(RColorBrewer)
library(GGally)
library(lubridate)

#Import data from CSV
data_campaign <- read.csv('P00000001-CA.csv',
                          sep = ",",
                          header = T,
                          row.names = NULL)
#Data wrangling
names(data_campaign) <- names(data_campaign)[-1]
data_campaign<-data_campaign[0:18]

#List of collumns
names(data_campaign)

#Subset the final candidates
finalists<-c('Clinton, Hillary Rodham','Trump, Donald J.')

#Subset the main candidates
main_candidates<- c(finalists, c(
  "Cruz, Rafael Edward 'Ted'",
  "Rubio, Marco",
  "Sanders, Bernard"))

#Highlight some election dates
#'08-11-16': General US presidential elections
#'28-07-16': Democrats convention
#'18-07-16': Republicans convention
#'3-05-16': Mr Sanders stop
#'15-03-16' : Mr Rubio stop

elections_dates<- c('08-11-16','28-07-16','18-07-16','3-05-16','15-03-16')

```

#### b) Summary of main variables

* cand_nm -> Name of the candidates

```{r echo=FALSE, message=FALSE, warning=FALSE}
summary(data_campaign$cand_nm)
```

* contb_receipt_amt -> Amount of the contribution

```{r echo=FALSE, message=FALSE, warning=FALSE}
summary(data_campaign$contb_receipt_amt)
```

* contb_receipt_city  -> City where the contribution was done

```{r echo=FALSE, message=FALSE, warning=FALSE}
summary(data_campaign$contbr_city)
```

* contbr_employer -> employer of contribution

```{r echo=FALSE, message=FALSE, warning=FALSE}
summary(data_campaign$contbr_employer)
```


#### c) Correlations

```{r echo=FALSE, message=FALSE, warning=FALSE}
wo.factors<-data_campaign
levels(wo.factors)

wo.factors$contb_receipt_amt<-as.numeric(wo.factors$contb_receipt_amt)
wo.factors$contbr_zip<-as.numeric(wo.factors$contbr_zip)
wo.factors$cand_id<-as.numeric(wo.factors$cand_id)
wo.factors$contb_receipt_dt<-dmy(
  wo.factors$contb_receipt_dt,
  locale = "english_us")

ggpairs(data = 
          wo.factors[,c("contb_receipt_amt", "contb_receipt_dt","cand_id")])


```






### III) First observations

#### a) Number of donators for each candidate

The first step is to assess the number of donators for each candidates.

The Y axis is in log10 scale as there is a huge difference between all candidates (Mr Gilmore vs Mrs Clinton)

```{r echo=FALSE, message=FALSE, warning=FALSE}

ggplot(data = subset(data_campaign,!is.na(cand_nm)), aes(x=cand_nm))+
  geom_bar(aes(fill=cand_nm), size=1, show.legend = F)+
  theme(axis.text.x=element_text(angle=90,vjust = 0))+
  labs(y = "Count of contributions", x="Candidates")+
  scale_y_log10()
  
```

The most interesting information we can get from this plot are:

* Huge difference between candidates (Mr Gilmore vs Mrs Clinton) 
* The relative low count of donations for Mr Trump (that was confirmed by a low score in popular vote: 32% vs 62% for Mrs Clinton ) that confirms that California was a "democrat"" state 


#### b) Amount of donations for each candidate

The next step is to assess the amount and no more the count of donations for each candidates.

The next scatter plot shows all donations for each candidats all along the campaign with the amount as Y axis.

```{r echo=FALSE, message=FALSE, warning=FALSE}


ggplot(data = 
         subset(data_campaign,
                cand_nm %in% main_candidates & contb_receipt_amt>0),
       aes(x=dmy(contb_receipt_dt, locale = "english_us"),y=contb_receipt_amt))+
  geom_jitter(alpha=1/4, position = 'identity', aes(color=cand_nm), shape=19)+
  scale_y_continuous(limits = c(0, 300))+
  theme(axis.text.x=element_text(angle=90,vjust = 0))+ 
  labs(y = "Amount of contributions ($)", x="Dates")+
  scale_x_date(date_breaks = "2 month" ,date_labels = "%b-%y",
               limits = c(dmy('01-04-15'), NA))+
  geom_vline(xintercept = dmy(elections_dates), show.legend = T)
  
```



Elections time line:

* 08-11-16: General US presidential elections, 
* 28-07-16: Democrats convention,
* 18-07-16: Republicans convention,
* 3-05-16: Mr Sanders stop,
* 15-03-16 : Mr Rubio stop



The most interesting information we can get from this plot are:

* The different steps (straight lines) corresponding to certain rounded values 
  + 10, 25, 50, 75, 100, 150, 200, 250, 300 
* The main colors depending on the date 
  + Mr Sanders from Oct. 15 to August 15, Mrs Clinton from Democrat convention to Election date and Mr trump at the very last stage of the campaign (ie.: after the general elections) 

On the next plot, I focused on the total amount of donations for the main candidates. 

```{r echo=FALSE, message=FALSE, warning=FALSE}

ggplot(data = subset(data_campaign,
                     (cand_nm %in% main_candidates ) &
                       contb_receipt_amt>1 &
                       !is.na(contb_receipt_amt)),
       aes(x=cand_nm,y=contb_receipt_amt))+
  geom_bar(stat='identity',aes(fill=cand_nm), size=1)+
  theme(axis.text.x=element_text(angle=90,vjust = 0))+
  labs(y = "Total amount of contributions ($)", x="Canditates")+
  scale_y_continuous()

```

The most interesting information we can get from this plot are:

* As for the count of donations, the total amount is far greater for Mrs Clinto than others 
* Low score for Mr Trump (lower than Mr Sanders) despite being elected in national elections at the end 

#### c) Amount of donations percontributors for main candidates

in the next plot, we can see the distribution of donation amounts for main candidates. 

```{r echo=FALSE, message=FALSE, warning=FALSE}

ggplot(data = subset(data_campaign,
                     cand_nm %in% main_candidates & contb_receipt_amt>0),
                      aes(x=cand_nm, y=contb_receipt_amt))+
  geom_boxplot(aes(fill=cand_nm))+
  theme(axis.text.x=element_text(angle=90,vjust = 0))+
  labs(y = "Contributions ($)", x="Canditates")+
  scale_y_continuous(limits = c(0, 150))


```

The most interesting information we can get from this box plot are:

* Mr Rubio contributors seem to spend more money than other candidates 
* Mrs Clinton contributors seem to spend less money than other candidates 

What could we assume about that observations:

* Are Mrs clinton's contributors  poorer than for other canditates? 
* Are Mrs Clinton's contibutors less committed than for other canditates? 

To answer this questions, we should have other informations like:

* Revenue of contributors 
* % satisfaction about the candidate for each contributors 

#### d) Focus on finalists 

The next table shows for both finalists the total sum and count of donations. 

```{r echo=FALSE, message=FALSE, warning=FALSE}
library (data.table)

# Create a data frame with count and sum of contrbutions for both finalists
donations_DT <- matrix(c(
  sum(subset(data_campaign,cand_nm==finalists[1])$contb_receipt_amt),
  length(subset(data_campaign,cand_nm==finalists[1])$contb_receipt_amt),
  sum(subset(data_campaign,cand_nm==finalists[2])$contb_receipt_amt),
  length(subset(data_campaign,cand_nm==finalists[2])$contb_receipt_amt)
  ),
  ncol=2,
  byrow=TRUE)
colnames(donations_DT) <- c("Sum", "n")
rownames(donations_DT) <- c('Clinton, Hillary Rodham','Trump, Donald J.')
donations_DT <- as.table(donations_DT)

# Display the data frame
donations_DT

```

We have a confirmation that Mrs Clinton receive much more money for her contributors. 

The next box plot confirms that mean donations are higher for Mr Trump than Mrs Clinton. 

```{r echo=FALSE, message=FALSE, warning=FALSE}
# Display a scatter plot for contributions amount for both finalists
ggplot(data = subset(data_campaign,!is.na(cand_nm) & contb_receipt_amt>0 &
                       (cand_nm == 'Trump, Donald J.' | 
                          cand_nm == 'Clinton, Hillary Rodham')),
                        aes(x=cand_nm, y=contb_receipt_amt))+
  geom_boxplot(aes(fill=cand_nm))+
  theme(axis.text.x=element_text(angle=90,vjust = 0))+
  labs(y = "Total amount of contributions ($)", x="Canditates")+
  scale_y_continuous(limits = c(0, 100))
```

### IV) Detailed plots


#### a) Evolution of donations for main candiates vs election time line

In the next chart, we can superpose the amount of donations all along the election. 

```{r echo=FALSE, message=FALSE, warning=FALSE}

# Display a line plot for contributions amount for main candidates
p1<-ggplot(data = subset(data_campaign,
                     cand_nm %in% main_candidates & contb_receipt_amt>0),
                      aes(x=dmy(contb_receipt_dt,
                                locale = "english_us"),
                          y=contb_receipt_amt))+
  geom_bar(stat='identity',show.legend = FALSE, aes(color=cand_nm))+
  theme(axis.text.x=element_text(angle=90,vjust = 0))+
  labs(y = "Total amount of contributions ($)", x="Dates")+
  scale_x_date(date_breaks = "2 month" ,date_labels = "%b-%y",
               limits = c(dmy('01-05-15'), dmy('01-12-16')))+
  geom_vline(xintercept = dmy(elections_dates), show.legend = T)+
  facet_wrap(~ cand_nm,
             ncol = 1, scale='free_y')+
  labs(title="Contributions during campaigin for both finalists")

p1
```

We can see that we have a clear correlation between donations and some milestone of the elections. 

For instance, Mr Sanders donations lowered after his capaign stop and completely vanished after Democrats convention. 

Another example is Mr Cruz with a clear stop of donations after his renonciation announcement.

#### b) Evolution of donations for finalists vs election time line

The same plot focused for finalists that confirm last analysis. 

```{r echo=FALSE, message=FALSE, warning=FALSE}

# Same line plot for both finalists
ggplot(data = subset(data_campaign,cand_nm %in% finalists & contb_receipt_amt>0),
       aes(x=dmy(contb_receipt_dt, locale = "english_us"),
           y=contb_receipt_amt))+
  geom_bar(stat='identity',aes(color=cand_nm, fill=cand_nm))+
  theme(axis.text.x=element_text(angle=90,vjust = 0))+
  labs(y = "Total amount of contributions ($)", x="Dates")+
  scale_x_date(date_breaks = "2 month" ,date_labels = "%b-%y",
               limits = c(dmy('01-04-15'), c(dmy('01-12-16'))))+
  geom_vline(xintercept = dmy(elections_dates))+
  facet_wrap(~cand_nm, ncol = 1, scale='free_y')

```

#### c) Contributor occupation for finalists

```{r echo=FALSE, message=FALSE, warning=FALSE}
library(dplyr)
library(tidyr)

# Create dataframe for finalists grouped by candidates and occupation
data_campaign.bycategory<- subset(data_campaign,cand_nm %in% finalists)  %>%
  group_by(cand_nm, contbr_occupation) %>%
  summarise(sum=sum(contb_receipt_amt),
            n=n()) %>%
  arrange(-n)

# Select TOP 13 of occupations
top_category<-head(subset(data_campaign.bycategory,
                          cand_nm %in% finalists),
                   n=13L)$contbr_occupation

# Remove outliers
top_category<- top_category[top_category!="INFORMATION REQUESTED"]
top_category<- top_category[top_category!=""]
top_category<- top_category[!is.na(top_category)]

# Change occupation to 'OTHER' if not part of TOP 13
data_campaign.bycategory<-transform(data_campaign.bycategory,
                                    contbr_occupation_new= 
                                      ifelse(
                                        contbr_occupation %in% top_category,
                                            as.character(contbr_occupation),
                                        as.character('OTHER')))

# Compute ratio field in the new data frame
data_campaign.bycategory<-transform(
  data_campaign.bycategory,
  ratio= ifelse(cand_nm=='Clinton, Hillary Rodham',
                n/donations_DT['Clinton, Hillary Rodham', 'n'],
                n/donations_DT['Trump, Donald J.', 'n']))

# Plot ratio of contributions for both finalits
ggplot(
  subset(
    data_campaign.bycategory,
    cand_nm %in% finalists),
  aes(
    x="",
    y=ratio,
    fill=contbr_occupation_new))+
  geom_bar(width = 1, stat = "identity")+
  facet_wrap(~cand_nm, ncol = 2)+
  labs(y = "Ratio", x="")+
  scale_color_brewer(type = 'div',
                     guide = guide_legend(
                       title = 'contbr_occupation',
                       reverse = T,
                       override.aes = list(alpha = 1, size = 1)))

```

We can see there are:

* More retired people donating for Mr Trump than for Mrs Clinton 
* More attorney, consutant, engineer, writer and teachers donating for Mrs Clinton than Mr Trump 

The same information is displayed in the following plot. 

```{r echo=FALSE, message=FALSE, warning=FALSE}

# Display occupation for finalists donators.
ggplot(subset(data_campaign.bycategory, cand_nm %in% finalists),
       aes(x=contbr_occupation_new, y=ratio))+
  geom_bar(
    width = 1,
    stat = "identity",
    aes(fill=contbr_occupation_new),
    show.legend = F)+
  theme(axis.text.x=element_text(angle=90,vjust = 0))+
  facet_wrap(~cand_nm, ncol = 1, scale='free_y')+
  labs(y = "Ratio", x="")

```

I propose to highlight the top 10 for both finalists.

```{r echo=FALSE, message=FALSE, warning=FALSE}

# TOP 10 contributors occupation for Clinton
TOP10_CLINTON <- head(
  subset(
    data_campaign.bycategory,
    cand_nm=='Clinton, Hillary Rodham'),
  n=11L)$contbr_occupation
# remove outlier
TOP10_CLINTON<-TOP10_CLINTON[TOP10_CLINTON!="INFORMATION REQUESTED"]

# TOP 10 contributors occupation for Clinton
TOP10_TRUMP <- head(
  subset(data_campaign.bycategory, cand_nm=='Trump, Donald J.'),
  n=11L)$contbr_occupation
# remove outlier
TOP10_TRUMP<-TOP10_TRUMP[TOP10_TRUMP!="INFORMATION REQUESTED"]

# TOP 10 contributors occupation for both finalists
TOP10_DT <- matrix(c(
  as.character(TOP10_CLINTON),
   as.character(TOP10_TRUMP)
  ),
  ncol=2)

colnames(TOP10_DT) <- c("Clinton, Hillary Rodham", "Trump, Donald J.")
rownames(TOP10_DT) <- c(seq(1,10,1))

TOP10_DT
```

We can notice that:

* Retired is the most common category for both candidates 
* Teacher, homemaker, unenployed, professor, consultant and writer are not present in Mr Trump Top 10 whereas in the Mrs Clinton Top10

#### d) Contributor city for finalists

```{r echo=FALSE, message=FALSE, warning=FALSE}
# Create dataframe for finalists grouped by candidates and city
data_campaign.bycity<- subset(data_campaign,cand_nm %in% finalists)  %>%
  group_by(cand_nm, contbr_city) %>%
  summarise(sum=sum(contb_receipt_amt),
            n=n()) %>%
  arrange(-n)

# Select TOP 20 of occupations
top_city<-head(subset(data_campaign.bycity,
                      cand_nm %in% finalists),
               n=30L)$contbr_city

# Display occupation for finalists donators.
p2<-ggplot(subset(data_campaign.bycity,
              cand_nm %in% finalists & contbr_city %in% top_city),
       aes(x=contbr_city, y=n, fill=contbr_city))+
  geom_bar(colour='black',  width = 1, stat = "identity",   show.legend = F)+
  theme(axis.text.x=element_text(angle=90,vjust = 0))+
  labs(y = "Count of contributions", x="", title="Contributions by City")+
  facet_wrap(~cand_nm, ncol = 1,
             scales = "free_y")

p2

```

We can notice that: 

* Bigger are the city more contributors we have. That is normal.
* On the other hand, we can notice that San Diego, San jose and Fresno are over represented in Mr Trump contributors
* In opposite way, Los Angeles and San Francisco are under represented (inn comparison with city populations)
* For Mrs Clinton, Biggest city as Los Angeles and San Franciso represents the main part of contributors. Oakland is big provider as well



The following plots will illustrate the same information in a more graphical view.

In a the first plot, we analyse Los Angeles map with number of conrtibutions.


```{r echo=FALSE, message=FALSE, warning=FALSE}

#Import data from CSV
zip.code <- read.csv('us_postal_codes.csv',
                          sep = ",",
                          header = T)

zip.code<-transform(zip.code,Zip.Code=as.character(Zip.Code))

# Create dataframe for finalists grouped by candidates and emplyers

data_campaign.byzip<-transform(data_campaign,
                               contbr_zip=substr(as.character(contbr_zip),1,5))


# Create dataframe for finalists grouped by candidates and emplyers
data_campaign.byzip<- subset(data_campaign.byzip,cand_nm %in% finalists)  %>%
  group_by(cand_nm, contbr_zip) %>%
  summarise(sum=sum(contb_receipt_amt),
            n=n()) %>%
  arrange(-n)

data_campaign.byzip.latlong <- data_campaign.byzip %>%
  full_join(zip.code, by = c("contbr_zip" = "Zip.Code" ))

data_campaign.byzip.latlong.losangeles <-
  subset(
    data_campaign.byzip.latlong,County=='Los Angeles' & cand_nm %in% finalists)


p3 <- qmplot(Longitude, Latitude,
             data = data_campaign.byzip.latlong.losangeles,
             size = n/400,
             color=cand_nm,show.legend=F,
             aes(alpha=I(0.5)))+
  facet_wrap(~cand_nm)+
  labs(title="Density of contributions for both finalist in Los Angeles")

p3
```

We can notice that Mrs Clinton contributors are concentrated in nort east of the city in comparison with Mr Trump contributors that are equally spred all over the city.

In the next plot, the same information are represented for San Diego that is the city where Mr Trump contributors were the most present (even if less numberous than Mrs Clinton's one)

```{r echo=FALSE, message=FALSE, warning=FALSE}
data_campaign.byzip.latlong.sandiego <-
  subset(
    data_campaign.byzip.latlong,County=='San Diego' &
      cand_nm %in% finalists)


qmplot(
  Longitude,
  Latitude, 
  data = data_campaign.byzip.latlong.sandiego,size = n/400,
  color=cand_nm, show.legend=F, 
  aes(alpha=I(0.5)))+
  facet_wrap(~cand_nm)+
  labs(title="Density of contributions for both finalist in San Diego")
```

On the last map, we have represented all California State.

Nothing special is to be noticed in this plot.

```{r echo=FALSE, message=FALSE, warning=FALSE}
data_campaign.byzip.latlong.california <-
  subset(
    data_campaign.byzip.latlong,State.Abbreviation=='CA' &
      cand_nm %in% finalists)


qmplot(Longitude,
       Latitude,
       data = data_campaign.byzip.latlong.california,
       size = n,
       color=cand_nm,
       show.legend=F,aes(alpha=I(0.1)))+
  facet_wrap(~cand_nm)+
  labs(title="Density of contributions for both finalist in California")
```


#### e) Contributor employer for finalists

I propose to look at contribution distribution per employers.

```{r echo=FALSE}

# Create dataframe for finalists grouped by candidates and emplyers
data_campaign.byemployer<- subset(data_campaign,cand_nm %in% finalists)  %>%
  group_by(cand_nm, contbr_employer) %>%
  summarise(sum=sum(contb_receipt_amt),
            n=n()) %>%
  arrange(-n)

# List the TOP 30 of employer
TOP_EMPLOYER<-head(subset(data_campaign.byemployer, cand_nm %in% finalists),
                   n=30L)$contbr_employer

#List of empoyer outliers
OUTLIERS_EMPLOYER<-c(
  'N/A',
  '',
  'RETIRED',
  'DISABLED',
  'HOMEMAKER',
  'SELF-EMPLOYED',
  'INFORMATION REQUESTED',
  'NOT EMPLOYED',
  'SELF EMPLOYED',
  'NONE' )

#Remove outliers
TOP_EMPLOYER<-TOP_EMPLOYER[!TOP_EMPLOYER %in% OUTLIERS_EMPLOYER]

# Display employers for finalists donators.
ggplot(subset(data_campaign.byemployer,
              cand_nm %in% finalists & contbr_employer %in% TOP_EMPLOYER),
       aes(x=contbr_employer,
           y=n, fill=contbr_employer))+
  geom_bar(colour='black',  width = 1, stat = "identity", show.legend = F)+
  theme(axis.text.x=element_text(angle=90,vjust = 0))+
  labs(y = "Count of contributions", x="")+
  facet_wrap(~cand_nm, ncol = 1, scales = "free_y")


```

We can notice that:

* State of California is one of biggest employer in California and so well represented for both finalists
* On the other hand, GAFA (Google, Apple and Facebook) are mainly represented in Mrs Clinton contirbutions (see following plot)

```{r echo=FALSE, message=FALSE, warning=FALSE}
GAFA <-c('GOOGLE', 'FACEBOOK', 'APPLE INC.')

# Display employers for finalists donators.

Total_sum_GAFA_CLINTON<-sum(subset(data_campaign.byemployer,
                                   cand_nm== "Clinton, Hillary Rodham" &
                                     contbr_employer %in% GAFA )['sum'])
Total_sum_GAFA_TRUMP<-sum(subset(data_campaign.byemployer,
                                 cand_nm== "Trump, Donald J." &
                                   contbr_employer %in% GAFA )['sum'])

df <- data.frame(
  Candidate = c("Clinton, Hillary Rodham", "Trump, Donald J."),
  Sum = c(Total_sum_GAFA_CLINTON, Total_sum_GAFA_TRUMP)
  )

ggplot(df, aes(x="", y=Sum, fill=Candidate))+
  geom_bar(width = 1, stat = "identity")+
  coord_polar("y", start=0)+
  labs(title = "Total of contributions per candidates for GAFA")
  

```

We can in the previous chart that GAFA contributors are for Mrs Clinton at an excessive rate (>99%)

### V) Summary

In this analysis, we have highligthed that:

* Mrs Clinton contributors are overepresented in California state in general and biggest city in particular (see following map)
  + especially in GAFA employers, biggest cities (SF, LA) and some occupation like attorneys or teachers
  + on the other hand, Mr Sanders was at the same level before his resignation at democrats convention
  
```{r echo=FALSE, message=FALSE, warning=FALSE}
p3
```



* Mr Trump contributors are under represented in California state (~15% of Mrs Clinton total)
  + Main city represented are: San Diego, Fresno and Beverly Hills (even if under Mrs Clinton rate) (see following plot)

```{r echo=FALSE, message=FALSE, warning=FALSE}
p2

```

  
* When we compare contribution vs campaign timeline we can a real correlation between donations and candidates status (running, suspended or stopped) (see following plot)

```{r echo=FALSE, message=FALSE, warning=FALSE}
p1
```



'08-11-16': General US presidential elections
'28-07-16': Democrats convention
'18-07-16': Republicans convention
'3-05-16': Mr Sanders stop
'15-03-16' : Mr Rubio stop

### VI) Reflection

With this limited dataset, we cannot conclude on several questions:

* How social, ethnic, religion or cultural environments impact the choice of candidate?
  + We could add information about revenue for example
  
* How satisfaction about selected candidate impact the contribution especially after conventions?
  + We could add a collumn in the dataframe with candidates polls

Regarding struggles and successes, I can highlith:

* It was very difficult to use some R api, I find documentation about R library non exhaustive and sometime frustrating

* Map visualization was quite easy to use

For me, R can be very powerful because we can display very interesting plots and charts in a short time frame. On the other hand, R becomes very difficult to use as soon as errors are raised as no complete documentations is available. Only help can come from Forums and it is why I did intensively once I got some big issues on my code.





